Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-11-12 paper/main.tex
section 6

\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
while (k > 0) {
  x=0;
  while (x < N && a[x] == 1) {
    @assert(na > 0);@
    a[x]=0; @na--;@
    tick(1);
    x++; }
  if (x < N) {
    a[x]=1; @na++;@
    tick(1); }
  k--;
}
\end{lstlisting}
\end{minipage}
\caption{Example of assisted bound derivation using logical state.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.  The bound derived automatically is
  $2|[0,k]| + |[0,na]|$.
  }
\label{fig:xmplincaux}
\end{figure}

\begin{figure}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant }
  \{ lg > \log_2(h - l) \} }$
bsearch(x,l,h@,lg@) {
  if (h-l > 1) {
    @assert(lg > 0);@
    m = l + (h-l)/2;
    @lg--;@ if (a[m]>x) h=m; else l=m;
    tick($M_\code{bsearch}$);
    l = bsearch(x,l,h@,lg@);
    tick($-M_\code{bsearch}$);
  } else return l;
}
\end{lstlisting}
\caption{Example of assisted bound derivation using logical state.
  We use the tick metric to model the stack consumption of the program.
  Before a function call to \code{bsearch}, a tick statement pays for
  the space used by stack frame of the function.  The space is recoverd
  after the call.  We write $\log_2(x)$ for the integer part of
  logarithm of $x$ in base 2.
  The bound derived automatically is $|[0,lg]|$.
  }
\label{fig:xmplbsaux}
\end{figure}

@@

While complete automation is often desirable, it cannot possibly
be universal since the problem is undecidable.  In this section
we present two techniques to derive complex resource bounds by
leveraging our automation.

When the resource bound of a program depends on the contents of the heap,
or is non-linear (e.g. logarithmic, exponential), we propose to
introduce \emph{logical state}.
This technique is inspired by XXX (speed) and consists in introducing
extra variables to the program that do not change its behavior, but help
the automatic tool to figure out a bound.

The systematic technique consists in the following steps. First,
a program $P$ that fails to be analyzed automatically is enriched
by extra variables $\vec x$ called logical variables to form a
program $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$
for the logical variables is shown to be such that:
$$
\forall n\, \sigma\, \sigma'.\,
  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
    {\implies}
  \exists n' {\le} n.\,
    (\sigma, P) \Downarrow_{n'} \sigma'.
$$
Since the annotated program and the original one are often
syntactically close, the proof of this result goes
by simple induction on the resource-aware evaluation judgement.
Third, using existing automation tools, a bound $B(\vec x)$ for
$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
$\vec X$, gives the final resource bound for the program $P$.

@@

On the other hand, it might be the case that even with annotations, a
resource bound cannot be found automatically.  With existing tools, this
situation requires a full rewrite of the program.  However, in our
framework, it is possible to write a manual proof and combine it with
bounds derived by the automatic analysis.  This interaction is made
formal using our Quantitative Hoare Logic.  This resource aware program
logic is an extension of the one presented in~\cite{veristack14} and
was originally designed to prove the soundness of our automatic
analysis.  To make it suitable for this end, we made it independent
of the resource used (e.g. stack usage, heap usage, clock ticks) and
implemented a smoother handling of the auxiliary state through a shallow
embedding in Coq.  Despite being first a theoretical tool, the logic can
be used directly as in~\cite{veristack14}.  Its judgements are inspired
by typical Hoare logics and have the shape
$$
\Delta \vdash \htriple P S Q
$$
where $\Delta$ is a set of hypothesis on program functions, and $P, Q$
are quantitative assertions, i.e. functions from the program states
to $\Qplusz \cup \{\infty\}$.  The semantics of triples and the rules
of the logic are described in further details in the extended version of
this paper~\cite{anon_extended}.  Because our automated tool produces
proofs of the resource bounds in the Quantitative Hoare Logic, it can
be used as an assistant when deriving a complex bound by hand in the
logic.  The interaction is guaranteed to be sound by the soundness
theorem of our Quantitative Hoare Logic.

\jan{ the following should go somewhere here:

Note that potential functions can access the whole program state, so
this framework can readily be used to derive bounds that depend on the
memory contents.
}

>>>>

\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant } \{ na = \#_1(a)} \}$
while (k > 0) {
  x=0;
  while (x < N && a[x] == 1) {
    @assert(na > 0);@
    a[x]=0; @na--;@
    tick(1); x++; }
  if (x < N) { a[x]=1; @na++;@ tick(1); }
  k--;
}
\end{lstlisting}
\end{minipage}
\caption{Assisted bound derivation using logical state.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 {\le} i {<} N {\land} a[i] {=} 1 \}$ and use the
  tick metric.  The derived bound  is
  $2|[0,k]| + |[0,na]|$.
  }
\label{fig:xmplincaux}
\end{figure}

\begin{figure}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant }
  \{ lg > \log_2(h - l) \} }$
bsearch(x,l,h@,lg@) {
  if (h-l > 1) {
    @assert(lg > 0);@
    m = l + (h-l)/2;
    @lg--;@ if (a[m]>x) h=m; else l=m;
    tick($M_\code{bsearch}$);
    l = bsearch(x,l,h@,lg@);
    tick($-M_\code{bsearch}$);
  } else return l;
}
\end{lstlisting}
\caption{Assisted bound derivation using logical state.
  % We use the tick metric to model the stack consumption of the program.
  % Before a function call to \code{bsearch}, a tick statement pays for
  % the space used by stack frame of the function.  The space is recoverd
  % after the call.
  We write $\log_2(x)$ for the integer part of
  logarithm of $x$ in base 2.
  The semi-automatically derived bound is $|[0,lg]|$.
  }
\label{fig:xmplbsaux}
\end{figure}

@@

While complete automation is desirable, it is not always possible
since the problem of bound derivation is undecidable.  In this section
we present a new technique to derive complex resource bounds
semi-automatically by leveraging our automation. Our goal is to
develop an interface between bound derivation and
established qualitative verification techniques.

When the resource bound of a program depends on the contents of the
heap, or is non-linear (e.g. logarithmic, exponential), we propose to
introduce a \emph{logical state} using \emph{auxiliary variables}.
Auxiliary variables do not change the behavior of the program and
guide $\toolname$ during bound derivation.

More precisely, the technique consists in the following steps. First,
a program $P$ that fails to be analyzed automatically is enriched by
auxiliary variables $\vec x$ to form a program
$P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$ for the
logical variables is selected so that:
$$
\forall n\, \sigma\, \sigma'.\,
  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
    {\implies}
  \exists n' {\le} n.\,
    (\sigma, P) \Downarrow_{n'} \sigma'.
$$
Since the annotated program and the original one are usually
syntactically close, the proof of this result goes
by simple induction on the resource-aware evaluation judgement.
Third, using existing automation tools, a bound $B(\vec x)$ for
$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
$\vec X$, gives the final resource bound for the program $P$.

@@

% On the other hand, it might be the case that even with annotations, a
% resource bound cannot be found automatically.  With existing tools, this
% situation requires a full rewrite of the program.  However, in our
% framework, it is possible to write a manual proof and combine it with
% bounds derived by the automatic analysis.  This interaction is made
% formal using our Quantitative Hoare Logic.  This resource aware program
% logic is an extension of the one presented in~\cite{veristack14} and
% was originally designed to prove the soundness of our automatic
% analysis.  To make it suitable for this end, we made it independent
% of the resource used (e.g. stack usage, heap usage, clock ticks) and
% implemented a smoother handling of the auxiliary state through a shallow
% embedding in Coq.  Despite being first a theoretical tool, the logic can
% be used directly as in~\cite{veristack14}.  Its judgements are inspired
% by typical Hoare logics and have the shape
% $$
% \Delta \vdash \htriple P S Q
% $$
% where $\Delta$ is a set of hypothesis on program functions, and $P, Q$
% are quantitative assertions, i.e. functions from the program states
% to $\Qplusz \cup \{\infty\}$.  The semantics of triples and the rules
% of the logic are described in further details in the extended version of
% this paper~\cite{anon_extended}.  Because our automated tool produces
% proofs of the resource bounds in the Quantitative Hoare Logic, it can
% be used as an assistant when deriving a complex bound by hand in the
% logic.  The interaction is guaranteed to be sound by the soundness
% theorem of our Quantitative Hoare Logic.
